---
title: "Clustering"
author: Aarushi Pandey ^[[Aarushi's Portfolio]()]
        Brandon Runyon ^[[Brandon's Portfolio]()]
        Zachary Canoot ^[[Zaiquiri's Portfolio](https://zaiquiriw.github.io/ml-portfolio/)]
        Gray Simpson ^[[Gray's Porfolio](https://ecclysium.github.io/MachineLearning_Portfolio/)]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    css: styles/bootstrap.css
    highlight: "kate"
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 80
---

# What is Our Data?

We like to think that a lot of people are going to use this database! When it
comes to clustering, the first use case is automatic recommendations for apps
like Netflix or Spotify. Our data, [Spotify most unpopular
songs](https://www.kaggle.com/datasets/estienneggx/spotify-unpopular-songs),
details different characteristics about 10,000+ of the top most disliked songs
on Spotify. Each track is an observation with characteristics about the music:

-   Danceability (In a range from 0 to 1)

-   Energy (In a range from 0 to 1)

-   The Key (Integers mapping to pitches the 12 pitches)

-   The Mode (Major and Minor although it be crazy if this could identify the
    likability of modes like mixolydian)

-   Loudness (Decimal in floating point) \* Speechiness (In a range from 0 to 1)

-   Acousticness (In a range from 0 to 1) \* Instrumentalness (In a range from 0
    to 1)

-   Liveness (Presence of audience in a range from 0 to 1)

-   [Valence](https://en.wikipedia.org/wiki/Valence_(psychology)) (Positiveness
    from 0 to 1)

As well as more basic statistics, like:

-   Tempo (BPM floating point)

-   Duration (Integer milliseconds)

-   Explicitness (True or False)

-   Popularity (I don't know how Spotify measures this integer, but 0 to 18)

And of course, the name of the track, it's id, and the artist. The beauty of all
this data is that it can all be scraped using the Spotify API, so any work with
this data can be extended to make custom tools that interact with Spotify. If we
apply clustering to this data, we will be able to form some groups of within the
data and understand what genres might be unpopular on Spotify.

We even have genre data that we can later use to analyze whether the
unsupervised clustering is getting close to the genres we saw. The current
hypothesis is that that is very unlikely due to the complexity of music, but we
shall see!

# Exploring Our Data

First we must load it in!

```{r}
spotify <- read.csv("data/unpopular_songs.csv")
head(spotify)
```

Looking good, although we would want to categorize the enumerated attributes key
and mode, as well as factor explicitness.

```{r}
# Make qualitative attributes into factors
spotify$key <- as.factor(spotify$key)
spotify$mode <- as.factor(spotify$mode)
spotify$explicit <- as.factor(spotify$explicit)

# Label the factors that are just enumerators
keys <- c("C", "C#", "D", "D#", "E", "F", "F#", "G", "G#", "A", "A#", "B")
modes <- c("Major", "Minor")
levels(spotify$mode) <- as.factor(modes)
levels(spotify$key) <- as.factor(keys)
str(spotify)
```

Now our data is easier to read! Note that I labeled the data major and minor,
while we don't actually know if 0 or 1 equals major or minor. I suspect a strong
bias towards the first factor being major. In the case of our clustering results
this doesn't matter!

## Statistics

While there might be slight correlations between various factors, you would
expect this data to be visually hard to see trends in.

```{r}
cor(spotify[])
```

## Graphics

# Clustering Algorithms

# Conclusion and Analysis
